---
title: "Transitional monitoring, fish, Baltic Sea"
output: 
  html_document: 
    code_fold: hide
    fig_caption: yes
    highlight: zenburn
    theme: journal
---

# Transitional monitoring, fish, Baltic Sea
#### Stressors: x1 = Mean nitrate, x2 = Max pH
#### Response: y =  Platichthys CPUE 

***
### Contents:
* 1 Data preparation
  + 1.1 Random Forest
  + 1.2 Dredge function
* 2 Linear regression modelling
  + 2.1 Data preparation
  + 2.2 Data transformation
  + 2.3 Stressor correlation
  + 2.4 Modelling
* 3 Model performance
* 4 Interaction plot

***

## 1 Data preparation

Load packages:
```{r, message=FALSE, warning=FALSE}
library("lme4")
library("MuMIn")
library("lmerTest")
library("nnet")
library("ordinal")
library("ape")
library("sp")
library("emdbook")
library("pacman")
library("car")
library("psych")
library("usdm")
library("PerformanceAnalytics")
library(randomForest)
library(pdp)
```

Load data:
```{r}
data <- read.csv("Data_prepared_Fish_year.csv", header = TRUE, sep = ";", dec = ".")
```

Display data summary:
```{r}
names(data)
str(data)
```

Set some important functions (for transformation and 3D plot):
```{r}
estimateBC = function(x){ 
  # function to estimate transformation parameters for continuous variable x require(car) 
  gamma = min(x, na.rm=T) - 0.001 # offset (min value minus a small number) 
  x = x - gamma # subtract gamma from x, so that it is strictly positive 
  lambda = powerTransform(x~1, family="bcPower")$lambda # estimate lambda of Box-Cox transformation... 
  xT = bcPower(x, lambda=lambda) # apply box-cox transform 
  xT.mean = mean(xT) # mean of transformed values, for centring 
  xT.sd = sd(xT) # sd of transformed values, for scaling 
  # return the transformation parameters 
  return(c(gamma=gamma, lambda=lambda, xT.mean=xT.mean, xT.sd=xT.sd)) 
}

applyBC = function(x, P=estimateBC(x)){ 
  # function to transform continuous variable x using transformation parameters P 
  require(car)
  gamma = P[1] 
  lambda = P[2] 
  xT.mean = P[3] 
  xT.sd = P[4] 
  xT = bcPower(x-gamma, lambda) # apply box-cox transform 
  xT = (xT-xT.mean)/xT.sd # centre and scale 
  return(xT) 
}

backBC = function(xT, P){ 
  # function to back transform transformed variable xT using transformation parameters P 
  gamma=P[1] 
  lambda=P[2] 
  xT.mean=P[3] 
  xT.sd=P[4]
  xT.unscaled = xT*xT.sd + xT.mean 
  x.original = exp(log(lambda*xT.unscaled + 1)/lambda) + gamma 
  return(x.original) 
}

# function to plot the response from a fitted model
interactionPlot = function(B, X1, X2, Y,
                           TP=list(P.x1=if(exists("P.x1")) P.x1 else NA,
                                   P.x2=if(exists("P.x2")) P.x2 else NA,
                                   P.y=if(exists("P.y")) P.y else NA,
                                   family="gaussian"),
                           responseLab="z", x1Lab="x1", x2Lab="x2") {
  # Function to plot interactions from fitted models.
  # B = vector of model fixed effect coeffcients
  # X1, X2 = vectors with values of the stressors (not transformed)
  # Y = vector with values of the response (not transformed)
  # TP = list of transformation parameters containing the following elements;
  # P.x1 = output from estimateBC() for x1, or NA if no transformation applied
  # P.x2 = output from estimateBC() for x2, or NA if no transformation applied
  # P.y = output from estimateBC() for y, or NA if no transformation applied
  # family = family of the generalised model. It should be one of
  # "gaussian" (for continuous response),
  # "poisson" (for count response),
  # "binomial" (for binary or ordered categorical response), or
  # NA (if you want to plot the response on the linear model scale)
  # responseLab = label for the response variables
  # x1Lab, x2Lab = labels for the stressor variables
  require(emdbook)
  if(is.numeric(X1) & is.numeric(X2)){ # X1 and X2 are both continuous
    myF <<- function(X1=X1, X2=X2, transPar=TP) {
      if(sum(is.na(transPar$P.x1))==0) X1 = applyBC(X1, transPar$P.x1)
      if(sum(is.na(transPar$P.x2))==0) X2 = applyBC(X2, transPar$P.x2)
      z = B[1] + B[2]*X1 + B[3]*X2 + B[4]*X1*X2
      if(!is.na(transPar$family)){
        if(transPar$family=="gaussian") z = backBC(z, transPar$P.y)
        if(transPar$family=="poisson") z = exp(z)
        if(transPar$family=="binomial") z = 1/(1+exp(-z))
      }
      return(z)
    }
    curve3d(myF, from=c(min(X1),min(X2)), to=c(max(X1),max(X2)),
            sys3d="image", col=colorRampPalette(c("blue","white","red"))(64),
            xlab=x1Lab, ylab=x2Lab, main=responseLab, varnames=c("X1","X2"))
    points(X1, X2, pch=21, bg="grey50")
    curve3d(myF, from=c(min(X1),min(X2)), to=c(max(X1),max(X2)), sys3d="contour",
            add=T, labcex=1, varnames=c("X1","X2"))
    box()
  }
  if(is.numeric(X1) & is.factor(X2)){ # X1 continuous, X2 is a factor
    myF <<- function(X1, X2=0, transPar=TP) {
      if(sum(is.na(TP$P.x1))==0) X1 = applyBC(X1, TP$P.x1)
      z = B[1] + B[2]*X1 + B[3]*X2 + B[4]*X1*X2
      if(!is.na(TP$family)){
        if(TP$family=="gaussian") z = backBC(z, TP$P.y)
        if(TP$family=="poisson") z = exp(z)
        if(TP$family=="binomial") z = 1/(1+exp(-z))
      }
      return(z)
    }
    curve(myF(X1=x, X2=0), from=min(X1), to=max(X1), col="blue", lwd=2, xlab=x1Lab, ylab=responseLab,
          ylim=if(is.numeric(y) & !is.na(TP$family)) range(y) else NULL)
    curve(myF(X1=x, X2=1), add=T, col="red", lwd=2, lty=2)
    rug(x1)
    legend("topright", legend=levels(X2), lty=1:2, lwd=2, col=c("blue","red"), title=x2Lab)
  }
  if(is.factor(X1) & is.numeric(X2)){ # X1 is a factor, X2 continuous
    myF <<- function(X1=0, X2, transPar=TP) {
      if(sum(is.na(TP$P.x2))==0) X2 = applyBC(X2, TP$P.x2)
      z = B[1] + B[2]*X1 + B[3]*X2 + B[4]*X1*X2
      if(!is.na(TP$family)){
        if(TP$family=="gaussian") z = backBC(z, TP$P.y)
        if(TP$family=="poisson") z = exp(z)
        if(TP$family=="binomial") z = 1/(1+exp(-z))
      }
      return(z)
    }
    curve(myF(X1=0, X2=x), from=min(X2), to=max(X2), col="blue", lwd=2, xlab=x2Lab,
          ylab=responseLab, ylim=if(is.numeric(y) & !is.na(TP$family)) range(y) else NULL)
    curve(myF(X1=1, X2=x), add=T, col="red", lwd=2, lty=2)
    rug(x2)
    legend("topright", legend=levels(X1), lty=1:2, lwd=2, col=c("blue","red"),
           title=x1Lab)
  }
  if(is.factor(X1) & is.factor(X2)){ # X1 is a factor, X2 is a factor
    z = c(B[1], B[1]+B[2], B[1]+B[3], B[1]+B[2]+B[3], sum(B))
    names(z) = c(paste(levels(X1)[1], levels(X2)[1], sep=" / "),
                 paste(levels(X1)[2], levels(X2)[1], sep=" / "),
                 paste(levels(X1)[1], levels(X2)[2], sep=" / "),
                 paste("E(", paste(levels(X1)[2], levels(X2)[2], sep=" + "), ")", sep=""),
                 paste(levels(X1)[2], levels(X2)[2], sep=" x "))
    #names(z) = paste(rep(levels(X1),2), rep(levels(X2),each=2), sep=" / ")
    if(!is.na(TP$family)){
      if(TP$family=="gaussian") z = backBC(z, TP$P.y)
      if(TP$family=="poisson") z = exp(z)
      if(TP$family=="binomial") z = 1/(1+exp(-z))
    }
    z2 = z - z[1] # z values minus the control
    par(mfrow=c(2,1))
    barplot(z, ylab=responseLab, xlab=paste(x1Lab,x2Lab,sep=" / "))
    abline(h=0)
    barplot(z2, ylab=paste(responseLab,"- control"), xlab=paste(x1Lab,x2Lab,sep=" / "))
    abline(h=0)
    par(mfrow=c(1,1))
  }
  return(NULL)
}
```

### 1.1 Select most important stressor variables using Random Forest:
```{r, class.source = 'fold-show'}
rfdata <- randomForest::randomForest(Platichthys.flesus.CPUE~., data=data[c(18,3:16)], na.action=na.roughfix, ntree=1000)
rfdata
varImpPlot(rfdata)

pdplot <- pdp::partial(rfdata, grid.resolution=30, pred.var = c("Nitrate.mean..umol.l.", "Salinity.mean..psu."), prob=T)
plotPartial(pdplot, prob=T)
```

### 1.2 Dredge to select 2 most important stressor variables
#### 1.2.1 Select variables x1 to x6, y and random effects:
```{r}
y  <-  data$Platichthys.flesus.CPUE # Response variable
x1 <-  data$Nitrate.mean..umol.l.
x2 <-  data$Phosphate.max..umol.l.
x3 <-  data$Salinity.mean..psu.
x4 <-  data$TN.max..umol.l.
x5 <-  data$Temperature.mean...C.
x6 <-  data$pH.max...


RE1 <- data$Region # Random Effect 1
RE1 <- as.factor(RE1)
RE2 <- data$Year # Random Effect 2
RE2 <- as.factor(RE2)
```


Remove cases with missing data:
```{r}
noMissing = !is.na(y) & !is.na(x1) & !is.na(x2) & !is.na(x3) & !is.na(x4)& !is.na(x5) & !is.na(x6) & !is.na(RE1)& !is.na(RE2) # TRUE for cases with no missing data, FALSE otherwise
y <-       y[noMissing]
x1 <-      x1[noMissing]
x2 <-      x2[noMissing]
x3 <-      x3[noMissing]
x4 <-      x4[noMissing]
x5 <-      x5[noMissing]
x6 <-      x6[noMissing]
RE1 <-     RE1[noMissing]
RE2 <-     RE2[noMissing]
```


Transform variables and check transformation:
```{r}
P.y = estimateBC(y)
yT = applyBC(y, P.y)
P.y

P.x1 = estimateBC(x1)
x1T = applyBC(x1, P.x1)
P.x1

P.x2 = estimateBC(x2)
x2T = applyBC(x2, P.x2)
P.x2

P.x3 = estimateBC(x3)
x3T = applyBC(x3, P.x3)
P.x3

P.x4 = estimateBC(x4)
x4T = applyBC(x4, P.x4)
P.x4

P.x5 = estimateBC(x5)
x5T = applyBC(x5, P.x5)
P.x5

P.x6 = estimateBC(x6)
x6T = applyBC(x6, P.x6)
P.x6

par(mfrow=c(2,4))
hist(y, br=12, col="gold3")
hist(x1, br=12, col="gold3")
hist(x2, br=12, col="gold3")
hist(x3, br=12, col="gold3")
hist(yT, br=12, col="gold3")
hist(x1T, br=12, col="gold3")
hist(x2T, br=12, col="gold3")
hist(x3T, br=12, col="gold3")

par(mfrow=c(2,3))
hist(x4, br=12, col="gold3")
hist(x5, br=12, col="gold3")
hist(x6, br=12, col="gold3")
hist(x4T, br=12, col="gold3")
hist(x5T, br=12, col="gold3")
hist(x6T, br=12, col="gold3")

shapiro.test(y)   # a high p-value = close to normal distribution
shapiro.test(yT)
shapiro.test(x1)   # a high p-value = close to normal distribution
shapiro.test(x1T)
shapiro.test(x2)   # a high p-value = close to normal distribution
shapiro.test(x2T)
shapiro.test(x3)   # a high p-value = close to normal distribution
shapiro.test(x3T)
shapiro.test(x4)   # a high p-value = close to normal distribution
shapiro.test(x4T)
shapiro.test(x5)   # a high p-value = close to normal distribution
shapiro.test(x5T)
shapiro.test(x6)   # a high p-value = close to normal distribution
shapiro.test(x6T)
```

Use dredge function to identify the two most important stressors incl. interaction:
```{r, class.source = 'fold-show'}
Mtot = lm(yT ~ (x1T+x2T+x3T+x4T+x5T+x6T)^2) # all main effects and 2-way interactions
summary(Mtot)
options(na.action = "na.fail") # needed for dredge
d = dredge(Mtot, trace=2, m.lim = c(0, 3)) # m.lim means consider only models with 0 to 3 fixed effects  
d # print model comparison table 
subset(d, delta <= 2) # print top ranked models (dAIC <= 2)
summary(get.models(d, 1)[[1]]) # summary of the top model (lowest AIC)
```

## 2 Linear regression modelling
### 2.1 Prepare: Load stressor and response variables:
```{r}
y  <-  data$Platichthys.flesus.CPUE # Response variable
x1 <-  data$Nitrate.mean..umol.l. # Stressor 1
x2 <-  data$pH.max... # Stressor 2

RE1 <- data$Region # Random Effect 1
RE1 <- as.factor(RE1)
RE2 <- data$Year # Random Effect 2
RE2 <- as.factor(RE2)
```


Remove cases with missing data:
```{r}
noMissing = !is.na(y) & !is.na(x1) & !is.na(x2) & !is.na(RE1)& !is.na(RE2) # TRUE for cases with no missing data, FALSE otherwise
y <-       y[noMissing]
x1 <-      x1[noMissing]
x2 <-      x2[noMissing]

RE1 <-     RE1[noMissing]
RE2 <-     RE2[noMissing]
```


Show sample size for the analysis and number of regions/years:
```{r, class.source = 'fold-show'}
length(y)
length(levels(RE1)) # print number of sites in the analysis
length(levels(RE2)) # print number of years in the analysis
```

Plot selected variables and check for outliers: 
```{r}
par(mfrow=c(1,3))
plot(x1,x2, pch=16, col=RE1)
plot(x1,y, pch=16, col=RE1)
plot(x2,y, pch=16, col=RE1)


boxplot(y,xlab="y", col="gold3")
boxplot(x1, xlab="x1", col="gold3")
boxplot(x2, xlab="x2", col="gold3")
par(mfrow=c(1,1))
```

### 2.2 Transform variables and check transformation:
```{r}
P.y = estimateBC(y)
yT = applyBC(y, P.y)
P.y

P.x1 = estimateBC(x1)
x1T = applyBC(x1, P.x1)
P.x1

P.x2 = estimateBC(x2)
x2T = applyBC(x2, P.x2)
P.x2


par(mfrow=c(2,3))
hist(y, br=12, col="gold3")
hist(x1, br=12, col="gold3")
hist(x2, br=12, col="gold3")

hist(yT, br=12, col="gold3")
hist(x1T, br=12, col="gold3")
hist(x2T, br=12, col="gold3")


shapiro.test(y)   # a high p-value = close to normal distribution
shapiro.test(yT)
shapiro.test(x1)   # a high p-value = close to normal distribution
shapiro.test(x1T)
shapiro.test(x2)   # a high p-value = close to normal distribution
shapiro.test(x2T)
```

### 2.3 Check correlation between stressors:
```{r, warning=FALSE}
dataT    <- cbind.data.frame(yT, x1T, x2T)
chart.Correlation(dataT[-c(4)], histogram=TRUE, pch=19, method = "spearman") # Spearman correlation should be < 0.7
```


Plot the main effects:
```{r, warning=FALSE}
par(mfrow=c(2,2))
plot(y ~ x1, pch=16, col=RE1)
plot(yT ~ x1, pch=16, col=RE1)

plot(y ~ x2, pch=16, col=RE1)
plot(yT ~ x2, pch=16, col=RE1)
par(mfrow=c(1,1))
```


### 2.3 Calculate GLMM including random effect:
```{r, warning=FALSE, class.source = 'fold-show'}
M = glm(yT ~ x1T*x2T)
r = residuals(M, type="response")
summary(M)
```

## 3 Model performance:
```{r, warning=FALSE, class.source = 'fold-show'}
r.squaredGLMM(M)

shapiro.test(r) # test for normality of residuals --> p value should be >0.05
par(mfrow=c(1,2))
hist(r, col="gold3") # histogram of residuals --> should look normally distributed
qqnorm(r, pch=16, col="gold3"); qqline(r) # plot residuals against normal quantiles --> points should be close to line
par(mfrow=c(1,1))

B = coef(M)
```

Check importance of interaction:
```{r, class.source = 'fold-show'}
(summary(M)$coef[,1]/summary(M)$coef[,2])["x1T:x2T"]    # z score of interaction (high posiive value --> important effect)
AIC(update(M, ~.-x1T:x2T)) - AIC(M)                     # change in AIC (high positive values --> important effect)
r.squaredGLMM(update(M, ~.-x1T:x2T)) - r.squaredGLMM(M) # change in R2 (negative --> important effect)
```

## 4 Draw interaction plot:
```{r, class.source = 'fold-show'}
myTP = list(P.x1=P.x1, P.x2=P.x2, P.y=P.y, family=NA)
interactionPlot(B=B, X1=x1, X2=x2, Y=y, TP=myTP, responseLab="z (model scale)", x1Lab="x1", x2Lab="x2")
```
