---
title: "Gradient analysis - Transitional monitoring, fish, Baltic Sea"
output: 
  html_document: 
    code_fold: hide
    fig_caption: yes
    highlight: zenburn
    theme: journal
---

***
### Contents
* 1 Data preparation
* 2 Data transformation
* 3 Gradient split
  + 3.1 Modelling of original dataset
  + 3.2 Create total, lower and upper gradient
* 4 Modelling
  + 4.1 Plot main effects & Check correlation between stressors
  + 4.2 GLMM of full and partial models
* 5 Loss in second stressor gradient length

***

## 1 Data preparation

Load packages:
```{r, message=FALSE, warning=FALSE}
require("lme4")
require("MuMIn")
require("lmerTest")
require("nnet")
require("ordinal")
require("ape")
require("sp")
require("car")
require("emdbook")
require("pacman")
p_load("psych")
require("usdm")
library("PerformanceAnalytics")
```

Load data and display summary:
```{r, warning=FALSE}
data <- read.csv("Case335.csv", header = TRUE, sep = ";", dec = ".")
str(data)
```

Define variables
```{r, warning=FALSE}
# define the response and explanatory variables ----
y <-   data$y # response
x1 <-  data$x1 # key stressor 1
x2 <-  data$x2 # key stressor 2
RE1 <- data$RandomEffect
```

## 2 Data transformation ----
```{r, message=FALSE, warning=FALSE}
estimateBC = function(x){ # function to estimate transformation parameters for continuous variable x
  gamma       = min(x, na.rm=T) - 0.001 # offset (min value minus a small number) 
  x           = x - gamma # subtract gamma from x, so that it is strictly positive 
  lambda      = powerTransform(x~1, family="bcPower")$lambda # estimate lambda of Box-Cox transformation... 
  xT          = bcPower(x, lambda=lambda) # apply box-cox transform 
  xT.mean     = mean(xT) # mean of transformed values, for centring 
  xT.sd       = sd(xT) # sd of transformed values, for scaling  
  # return the transformation parameters 
  return(c(gamma=gamma, lambda=lambda, xT.mean=xT.mean, xT.sd=xT.sd)) }

applyBC = function(x, P=estimateBC(x)){ # function to transform continuous variable x using transformation parameters P 
  gamma = P[1] 
  lambda = P[2] 
  xT.mean = P[3] 
  xT.sd = P[4] 
  xT = bcPower(x-gamma, lambda) # apply box-cox transform 
  xT = (xT-xT.mean)/xT.sd # centre and scale 
  return(xT) }

P.y = estimateBC(y)
yT = applyBC(y, P.y)
P.y

P.x1 = estimateBC(x1)
x1T = applyBC(x1, P.x1)
P.x1

P.x2 = estimateBC(x2)
x2T = applyBC(x2, P.x2)
P.x2
```

## 3 Gradient split
### 3.1 Modelling of original dataset
```{r, warning=FALSE}
M = lmer(yT ~ x1T*x2T + (1|RE1), REML = FALSE)
summary(M)
r.squaredGLMM(M)
```

### 3.2 Create total, lower and upper gradient
```{r, warning=FALSE}
originaldf    <- cbind.data.frame(yT, x1T, x2T, RE1)

originaldf   <- originaldf[order(originaldf$x1T),]
toDelete <- seq(1, nrow(originaldf), 2)
totdf <- originaldf[ toDelete ,]

lowdf    <- originaldf[originaldf$x1T <= median(originaldf$x1T),]
highdf   <- originaldf[originaldf$x1T >  median(originaldf$x1T),]

str(totdf)
str(lowdf)
str(highdf)
```

Compare full and partial datasets:
```{r, warning=FALSE}
totsum <- cbind("Total gradient", length(totdf$yT), min(totdf$yT), max(totdf$yT),  min(totdf$x1T), max(totdf$x1T), min(totdf$x2T), max(totdf$x2T))
lowsum <- cbind("Low gradient", length(lowdf$yT), min(lowdf$yT), max(lowdf$yT),  min(lowdf$x1T), max(lowdf$x1T), min(lowdf$x2T), max(lowdf$x2T))
highsum <- cbind("High gradient", length(highdf$yT), min(highdf$yT), max(highdf$yT),  min(highdf$x1T), max(highdf$x1T), min(highdf$x2T), max(highdf$x2T))

dataset.comp <- setNames(rbind.data.frame(totsum, lowsum, highsum), c("Dataset", "sample.size", "min.y", "max.y", "min.x1", "max.x1", "min.x2", "max.x2"))
dataset.comp
```

## 4 Modelling

### 4.1 Plot main effects & Check correlation between stressors
```{r, warning=FALSE}
chart.Correlation(totdf[-c(4,5)], histogram=TRUE, pch=19, method = "spearman")
chart.Correlation(lowdf[-c(4,5)], histogram=TRUE, pch=19, method = "spearman")
chart.Correlation(highdf[-c(4,5)], histogram=TRUE, pch=19, method = "spearman")
```

### 4.2 GLMM of full and partial gradients
```{r, warning=FALSE}
Mtot <-  lmer(data=totdf, yT ~ x1T*x2T + (1|RE1), REML = FALSE)
summary(Mtot)
R2tot <- r.squaredGLMM(Mtot)
R2tot

Mlow <-  lmer(data=lowdf, yT ~ x1T*x2T + (1|RE1), REML = FALSE)
summary(Mlow)
R2low <- r.squaredGLMM(Mlow)
R2low

Mhigh <-  lmer(data=highdf, yT ~ x1T*x2T + (1|RE1), REML = FALSE)
summary(Mhigh)
R2high <- r.squaredGLMM(Mhigh)
R2high
```

look at residuals:
```{r, warning=FALSE}
rtot <-  residuals(Mtot, type="response")
shapiro.test(rtot) # test for normality of residuals --> p value should be high
par(mfrow=c(1,2))
hist(rtot) # histogram of residuals
qqnorm(rtot); qqline(rtot) # plot residuals against normal quantiles
par(mfrow=c(1,1))
plot(fitted(Mtot), rtot) # fitted vs residuals plot
Ptot <- cor.test(rtot, fitted(Mtot)) # test correlation between residuals and fitted
Ptot$estimate^2

rlow <-  residuals(Mlow, type="response")
shapiro.test(rlow) # test for normality of residuals --> p value should be high
par(mfrow=c(1,2))
hist(rlow) # histogram of residuals
qqnorm(rlow); qqline(rlow) # plot residuals against normal quantiles
par(mfrow=c(1,1))
plot(fitted(Mlow), rlow) # fitted vs residuals plot
Plow <- cor.test(rlow, fitted(Mlow)) # test correlation between residuals and fitted
Plow$estimate^2

rhigh <-  residuals(Mhigh, type="response")
shapiro.test(rhigh) # test for normality of residuals --> p value should be high
par(mfrow=c(1,2))
hist(rhigh) # histogram of residuals
qqnorm(rhigh); qqline(rhigh) # plot residuals against normal quantiles
par(mfrow=c(1,1))
plot(fitted(Mhigh), rhigh) # fitted vs residuals plot
Phigh <- cor.test(rhigh, fitted(Mhigh)) # test correlation between residuals and fitted
Phigh$estimate^2
```

## 5 Loss in second stressor gradient length
```{r, warning=FALSE}
minmax <- function(x, y){(max(x)-min(x))/(max(y)-min(y))}
minmax(lowdf$x1T, totdf$x1T)
minmax(highdf$x1T, totdf$x1T)
```